Adversarial deep neural networks remove nonlinear batch effects from gene-expression data

### 2019-07-19 manuscript.tex
Add description of competing methods in comments.
Updated manuscript

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/adjuster_workflow.pdf}
	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) may affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

@@

\begin{table}
	\centering
	\caption[Dataset information]{
		\textbf{Information} about each dataset used to compare algorithm performance.
	}
	\input{figures/final/datasets.tex}
	\label{tab:datasets}
\end{table}

@@

The Cancer Genome Atlas (TCGA) produced RNA-Sequencing data for tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
In that study, we observed that cancer type had a strong confounding effect on our analysis, in which sought to identify patterns spanning many cancer types. We adjusted for this effect using ComBat.
However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
Here, we used the same version of TCGA dataset (available at \url{https://osf.io/7xjdn/}),
which has expression values for 9,365 samples across 25 cancer types.

@@

We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
[@The defaults in version 0.19.1 of sklearn are n_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
		Lower MSE represents that the adjuster has more faithfully reproduced the input data.
		MSE for unadjusted data will always be 0 because the input data is identical to the output data.
		Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mse}.
	}
	\label{fig:mse}
\end{figure}
\begin{table}
	\input{figures/final/mse.tex}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
		Lower MSE indicates that the output has changed less from the input.
		See also \figurename{} \ref{fig:mse}.
	}
	\label{tab:mse}
\end{table}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) between different batches.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mmd}.
	}
	\label{fig:mmd}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/mmd.tex}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		See also \figurename{} \ref{fig:mmd}.
	}
	\label{tab:mmd}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/batch.tex}
	\caption[Batch classification accuracy]{
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times. A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{figures/final/true_class.tex}
	\caption[True class classification accuracy]{
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also \figurename{} \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}

>>>>

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/static/adjuster_workflow.pdf}
	\caption[Batch adjustment justification and steps]{\textbf{Batch adjustment justification and steps.}
	\begin{enumerate*}[(a)]
		\item When measurements are collected from a sample ($X$), systemic effects ($H$) may affect the measurements.
		\item If data from the same sample $X$ is measured under two different conditions, $H_1$ and $H_2$, we may obtain slightly different measurements.
		\item In order to normalize batches of data relative to one another, we first estimate the effect of the hidden variables based on differences in measurements between batches.
		\item Second, we remove the estimated effects in order to normalize the batches relative to one another.
	\end{enumerate*}
	}
	\label{fig:workflow}
\end{figure}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/static/network.pdf}
	\caption[Network architecture of Confounded]{\textbf{Network architecture of Confounded.}
	Data with batch effects (represented by different colors) are input into an autoencoder.
	The output of the autoencoder is classified by a discriminator network based on batch.
	The autoencoder is then penalized based on the success of the discriminator.
	Over time, the autoencoder learns to output a faithful representation of the data without signal due to batch.}
	\label{fig:network}
\end{figure}

@@

% ----- Brief explanation of the other methods -----
% 1.
% The Shaham lab has two method that we reference up here^, one from 2017 , and one from 2018.
% The 2017 one is the distribution-matching residual net.
% It arbitrarily designates one batch as the "target" and the other batch as the "source".
% It calculates MMD with 3 kernels and averages them, and uses them as the loss function.
% Over training, the source's distribution (according to MMD) should be aligned to the target's distribution.
% In our testing, this was SUPER memory-intensive.
% I think this is the practical reason they reduced their dataset to the first 37 principal components, but their explanation in the paper is that the people who generated the data said the first 37 components contain most of the variation.
% 2.
% The 2018 Shaham paper is about using an adversarial autoencoder to make a batch-free encoding of expression data.
% They have a variational autoencoder with a shared encoder, but separate decoders for each of the 2 batches.
% They say that the decoders output data that has batch info in it but that the encoding layer doesn't have batch info.
% Their encoding layer is only size 15-20, so this is essentially a dimensionality reduction technique--the "batch free encoding" they're producing doesn't have values for each gene.
% 3.
% I don't know how worth-mentioning the Upadhyay paper is, but they basically replicate the DMResNet paper and replace the MMD loss with a discriminator loss like we do.
% They use the Shaham datasets, and I think their code, which is either not publicly available or is just really hard to find, probably has the same caveats as the Shaham code.
% Most particularly, they only account for 2 batches.
% I might have misunderstood their approach though because the English is a little difficult to interpret.

@@

\begin{table}
	\centering
	\caption[Dataset information]{
		\textbf{Information} about each dataset used to compare algorithm performance.
	}
	\input{tables/static/datasets.tex}
	\label{tab:datasets}
\end{table}

@@

The Cancer Genome Atlas (TCGA) produced RNA-Sequencing data for tumors across many cancer types \cite{the_cancer_genome_atlas_research_network_cancer_2013}.
In a previous study \cite{dayton_classifying_2017-1}, we classified this dataset based on the presence or absence of mutations in several known cancer genes.
In that study, we observed that cancer type had a strong confounding effect on our analysis, in which sought to identify patterns spanning many cancer types. We adjusted for this effect using ComBat.
However, we found that a strong nonlinear signal could still be identified by the Random Forests algorithm after adjustment.
Here, we used the same version of TCGA dataset (available at \url{https://osf.io/7xjdn/}),
which has expression values for 9,365 samples across 25 cancer types.
%TODO: Change this to 8744 samples (after averaging across replicates and limiting to patients that had all types of data), 23,386 genes and 24 cancer types?
%TODO: Mention that we used all genes rather than 300?

@@

%TODO: Update the algorithms listed here.
We used four classification algorithms from the scikit-learn (version 0.19.1) library \cite{pedregosa_scikit-learn_2011}: Naive Bayes \citep{maron_automatic_1961}, Random Forests \citep{tin_kam_ho_random_1995}, k-Nearest Neighbors \citep{fix_discriminatory_1951}, and Support Vector Machines (SVM) \citep{cortes_support-vector_1995} with a radial-basis kernel.
In most cases, we used the default hyperparameters, with the exceptions of using Random Forests with \texttt{n\_estimators=10} and using SVM with \texttt{kernel="rbf"}.
[@The defaults in version 0.19.1 of sklearn are n\_estimators=10 for RF and kernel="rbf" for SVM. Did you use these defaults or something different?@]

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mse.pdf}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error} (MSE) between the data prior to and after adjustment with various algorithms.
		Lower MSE represents that the adjuster has more faithfully reproduced the input data.
		MSE for unadjusted data will always be 0 because the input data is identical to the output data.
		Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mse}.
	}
	\label{fig:mse}
\end{figure}
\begin{table}
	\input{tables/static/mse.tex}
	\caption[Mean squared error (MSE)]{
		\textbf{Mean squared error (MSE)} of the unadjusted input data compared to the data output by the given adjusters.
		Lower MSE indicates that the output has changed less from the input.
		See also \figurename{} \ref{fig:mse}.
	}
	\label{tab:mse}
\end{table}
\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/mmd.pdf}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) between different batches.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		In each case, Confounded usually performs better than the scale adjuster and somewhat worse than ComBat when measuring MSE.
		See also Table \ref{tab:mmd}.
	}
	\label{fig:mmd}
\end{figure}
\begin{table}
	\centering
	\input{tables/static/mmd.tex}
	\caption[Maximum mean discrepancy (MMD)]{
		\textbf{Maximum mean discrepancy} (MMD) comparing the distributions of the batches to each other after a given adjustment.
		Lower MMD indicates that the distributions of the different batches are more similar.
		In cases with more than two batches, MMD is computed pairwise between each batch and averaged.
		See also \figurename{} \ref{fig:mmd}.
	}
	\label{tab:mmd}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/batch_accuracy.pdf}
	\caption[Batch classification accuracy]{\textbf{Batch classification accuracy} from 4-fold cross-validation repeated 3 times for several classifiers.
	Lower batch accuracy indicates that more batch-related signal has been removed and therefore indicates better performance.
	Confounded's performance is similar to ComBat's for the smaller datasets and is improved for the larger datasets.
	See also Table \ref{tab:batch}.}
	\label{fig:batch}
\end{figure}
\begin{table}
	\centering
	\input{tables/static/batch.tex}
	\caption[Batch classification accuracy]{
		\textbf{Batch classification accuracy} for several datasets and adjusters.
		The ideal batch adjuster would completely remove all signal due to batch and would therefore \textit{decrease} batch classification accuracy to around the baseline for all classifiers.
		See also \figurename{} \ref{fig:batch}.
	}
	\label{tab:batch}
\end{table}

@@

\begin{figure}
	\centering
	\includegraphics[width=\columnwidth]{figures/final/true_class_accuracy.pdf}
	\caption[True class classification accuracy]{\textbf{True class classification accuracy} for several datasets and adjusters with 4-fold cross-validation repeated 3 times. A higher accuracy after adjustment is desired because it represents that the adjuster has not destroyed the true class signal.
	See also Table \ref{tab:true_class}.}
	\label{fig:true_class}
\end{figure}
\begin{table}
	\centering
	\input{tables/static/true_class.tex}
	\caption[True class classification accuracy]{
		\textbf{True class classification accuracy} for several datasets and adjusters.
		After adjustment by the ideal batch adjuster, all true class signal should be preserved, and all classifiers should therefore have the same accuracy in predicting true class before and after adjustment.
		See also \figurename{} \ref{fig:true_class}.
	}
	\label{tab:true_class}
\end{table}
